### 2021-02-28 newvariant.tex
done for now?

Characterizing epidemic dynamics of new SARS-CoV-2 variants is critical to predicting the future course of the pandemic.
Here, we explore how assumptions about the relative strength (i.e., ratio of reproduction numbers $\RR$) and relative speed (i.e., differences in growth rates $r$) of the new variant affect conclusions about each other, given the possibility that the new variant has longer generation intervals.
When relative speed---which corresponds to the growth rate of the logistic curve that models the proportion of the new variant---is held constant, higher growth rate of the original strain leads to higher relative strength.
When relative strength is held constant, however, higher reproduction number of the original strain leads to smaller relative speed.
Therefore, intervention strategies that reduce $\RR$ can cause the variant to replace the original strain faster.
Neglecting differences in generation intervals lead to opposite conclusions about the relationship between relative strength and speed, biasing their estimates.
We suggest that estimates of increased transmissibility of new variants should be re-assessed in the light of these findings.

@@

Many studies have tried to estimate the strength of the novel variant from the observed speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
On the other hand, some studies have assumed a value of the strength of the novel variant and tried to predict future spread (therefore, the speed) of the variant \citep{davies2021estimated}.
While both approaches are reasonable, they can lead to different conclusions.
In many contexts, previous studies have shown that inferences about disease dynamics (e.g., effectiveness of intervention strategy) can changes depending on whether $\RR$ or $r$ is held constant \citep{dushoff2020speed}.
Such inference also depends on assumptions about the underlying generation-interval distribution \citep{roberts2007model,svensson2007note,wallinga2007generation}.
While many studies assume that the generation-interval distributions do not differ between the existing strain and the variant, recent evidence suggests that the new variant has longer infectious period, leading to longer generation intervals.

Here, we explore how different assumptions about the new variant affects estimates of its strength and speed.
We ask how the estimates of relative strength (i.e., the ratio between the strength of the existing strain and the variant) and speed (i.e., the difference between the speed of the existing strain and the variant) depend on each other.
We then show how neglecting differences in the generation-interval distributions can lead to biased estimates.

@@

While relative speed is easier to measure, we often do not expect $\delta$ to be constant---this can depend on epidemic and demographic factors and can vary across time and space.
Instead, assuming a constant relative strength $\theta$ better matches biological hypotheses about the new variant (e.g., increased transmission and immune escape).
Several studies have tried to predict the future spread of the new variant, including when it will replace the wild type (therefore, its relative speed $\delta$).
As our baseline scenario we assume $\theta = 1.8$ and ask how assumptions about the generation-interval distribution affect predictions of $\delta$.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{relspeed.pdf}
\caption{
\textbf{Relative speed of the new variant given assumed relative strength.}
(A) True relative speed of the new variant $\delta$ given strength of the wild type $\RR_w$ and the relative strength $\theta$.
The dashed contour line corresponds to $\delta = 0$.
True relative speed of the new variant $\delta$ is calculated using $\bar{G}_w = 5\,\textrm{days}$, $\bar{G}_v = 8\,\textrm{days}$, and $\kappa = 1/5$. 
(B) Estimated relative speed of the new variant $\hat{\delta}$ given strength of the wild type $\RR_w$ and the relative strength $\theta$ assuming identical generation-interval distributions.
(C) Biases in the estimates of the relative speed of the new variant $\hat{\delta} - \delta$ given strength of the wild type $\RR_w$ and the relative strength $\theta$.
The dashed contour line corresponds to $\hat{\delta} = \delta$.
}
\label{fig:relspeed}
\end{figure}

@@

Here, we explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
We define relative strength as the ratio between reproduction numbers ($\theta=\RR_v/\RR_w$) and relative speed as the difference between growth rates ($\delta=r_v-r_w$) of the variant and the wild type.
When the generation-interval distributions are asumed to be identical, we find that $\hat{\theta}$ decreases with $$

When relative speed $\delta$ is held constant, faster growth of the existing strain leads to larger relative strength.
When relative strength $\theta$ is held constant, higher strength of the existing strain leads to smaller relative speed.
Failing to account for the differences in the generation-interval distributions lead to opposite conclusions, leading to biased estimates of $\delta$ and $\theta$.

We argue that both perspectives (i.e., whether we hold relative strength or speed constant) are useful in understanding the dynamics of the new variant.
Early in the spread, we typically want to fix the relative speed because we can directly observe it.
As more information about the transmission and immunity profiles of the new variant becomes available, it is better to fix the relative strength as it better mathces biological assumptions.
We suggest researchers to be mindful about which quantity they hold constant and how their assumptions lead to their conclusions.

Our simple framework further reveals counterintuitive, but important, insight.
In general, one would expect that stronger intervention strategies is required in order to slow down the replacement of the original strain by the new variant.
However, this is only true if both have the same mean generation intervals (\fref{relspeed}B).
If the new variant has longer generation intervals than the original strain, imposing strong intervention (therefore reducing both $\RR_w$ and $\RR_v$ by a constant amount) results in a higher relative speed $\delta$, and therefore faster replacement of the original strain (\fref{relspeed}A).
While it is still necessary to impose stronger intervention to control the epidemic if the new variant has higher $\RR$, it may have unexpected consequences.

Since the beginning of the SARS-CoV-2 pandemic, the importance of understanding the role of generation-interval distributions in its spread has been constantly emphasized.
Yet, we still feel that generation intervals are under-appreciated, often viewed as a leeway to estimating $\RR$.
Future studies should consider the possibility that the new variant might have longer generation intervals and evaluate how assumptions about generation intervals affect their conclusions.

>>>>

Characterizing epidemic dynamics of new SARS-CoV-2 variants is critical to predicting the future course of the pandemic.
Here, we explore how assumptions about the relative strength (i.e., ratio of reproduction numbers $\RR$) and relative speed (i.e., differences in growth rates $r$) of the new variant affect conclusions about each other.
We show that failing to account for the possibility that the new variant has longer generation intervals can lead to considerable biases in the inference of the spread of the new variant.
For example, assuming a mean generation interval of 7 days for the new variant and 5 days for the wild type suggest that 1.1--1.4 fold biases in the degree of increased transmissibility of the new variant is plausible.
We suggest that estimates of increased transmissibility of new variants should be re-assessed in the light of these findings.

@@

Many studies have tried to estimate the strength of the novel variant from the observed speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
On the other hand, some studies have assumed a value of the strength of the novel variant and tried to predict future spread (therefore, the speed) of the variant \citep{davies2021estimated}.
While both approaches are reasonable, they can lead to different conclusions.
In many contexts, previous studies have shown that inferences about disease dynamics (e.g., effectiveness of intervention strategy) can changes depending on whether $\RR$ or $r$ is held constant \citep{dushoff2020speed}.

Such inference also depends on assumptions about the underlying generation-interval distribution \citep{roberts2007model,svensson2007note,wallinga2007generation}.
Most studies have assumed that the generation-interval distributions do not differ between the existing strain and the variant, but recent evidence suggests that the new variant may have longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
Longer duration of infection provides opportunity to transmit for a longer period of time, meaning that the generation intervals of the new variant is likely to be longer than that of the wild type.
However, there is much uncertainty in how much longer generation intervals might be because generation intervals also depend on other factors that affect transmission such as awareness-driven behavior (e.g., faster self-isolation after symptom onset leading to shorter generation intervals).

Here, we explore how different assumptions about the new variant affects estimates of its strength and speed given the possibility that it might have longer generation intervals.
We ask how the estimates of relative strength (i.e., the ratio between the strength of the existing strain and the variant) and speed (i.e., the difference between the speed of the existing strain and the variant) depend on each other.
We then show how neglecting differences in the generation-interval distributions can lead to biased estimates.

@@

While relative speed is easier to measure, we often do not expect $\delta$ to be constant---this can depend on epidemic and demographic factors and can vary across time and space.
Instead, assuming a constant relative strength $\theta$ better matches biological hypotheses about the new variant (e.g., increased transmission and immune escape).
Several studies have tried to predict the future spread of the new variant, including when it will replace the wild type (therefore, its relative speed $\delta$).
As our baseline scenario we assume $\theta = 1.8$ and ask how assumptions about the generation-interval distribution affect predictions of $\delta$---while this is considerably higher than most previous estimates of around 1.5 \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}, we chose this value as it better reflects the possibility that the new variant may have longer generation intervals (\fref{relstrength}).

\begin{figure}[!th]
\includegraphics[width=\textwidth]{relspeed.pdf}
\caption{
\textbf{Relative speed of the new variant given assumed relative strength.}
(A) Estimated relative speed of the new variant $\hat{\delta}$ given strength of the wild type $\RR_w$ assuming identical generation-interval distributions.
(B) True relative speed of the new variant $\delta$ given strength of the wild type $\RR_w$ accounting for longer generation intervals of the new variant.
(C) Difference between the true and the estimated relative speed of the new variant $\delta - \hat{\delta}$.
True relative speed of the new variant $\delta$ is calculated using $\theta=1.8$, $\bar{G}_w = 5\,\textrm{days}$, and $\bar{G}_v = 7\,\textrm{days}$.
}
\label{fig:relspeed}
\end{figure}

@@

Here, we explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
We define relative strength as the ratio between reproduction numbers ($\theta=\RR_v/\RR_w$) and relative speed as the difference between growth rates ($\delta=r_v-r_w$) of the variant and the wild type.
When the generation-interval distributions are asumed to be identical, we find that estimates of $\hat{\theta}$ decreases with $r_w$ whereas $\hat{\delta}$ increase with $\RR_w$.
When we account for longer generation intervals of the new variant, these relationships can change directions depending on the effective dispersion in generation intervals, which we characterized in terms of the squared coefficient of variation of the gamma distribution $\kappa$.
Therefore, even a modest differences in the mean generation intervals (2 days, in this case) can lead to considerable biases in the inference---we find that 1.1--1.4 fold under-estimation in the relative strength of the new variant is plausible.

We argue that both perspectives (i.e., whether we hold relative strength or speed constant) are useful in understanding the dynamics of the new variant.
Early in the spread, we typically want to fix the relative speed because we can directly observe it.
As more information about the transmission and immunity profiles of the new variant becomes available, it is better to fix the relative strength as it better matches biological mechanisms of its spread.
We suggest researchers to be mindful about which quantity they hold constant and how their assumptions lead to their conclusions.

Our simple framework further reveals counterintuitive, but important, insight.
In general, one would expect that stronger intervention strategies is required in order to slow down the replacement of the original strain by the new variant.
However, this is only true if both have the same mean generation intervals (\fref{relspeed}B).
If the new variant has longer generation intervals than the original strain, imposing strong intervention (therefore reducing both $\RR_w$ and $\RR_v$ by a constant amount) can result in a higher relative speed $\delta$, and therefore faster replacement of the original strain (\fref{relspeed}B).
This is not necesarily a bad news as the absolute speed ($r_v$) of the new variant still decreases as $\RR_v$ decreases.
These effects should be taken into consideration for accurate projection of the spread of the new variant.

Even though SARS-CoV-2 has been spreading for more than a year, there is still large uncertainty in its generation intervals.
Observational studies typically focus on serial intervals (i.e., time between symptom onset of the infector and the infectee; \cite{svensson2007note}) because they are easier to measure \citep{griffin2020rapid}, but they are subject to dynamical biases that can be difficult to tease apart \citep{park2021forward}.
A few studies have tried to estimate the generation-interval distribution from serial intervals with means ranging between 3--6 days and squared coefficient of variations ranging between 0.1--0.6 \citep{ferretti2020quantifying,Ferretti2020timing,ganyani2020estimating,knight2020estimating}; 
however, most estimates neglect dynamical biases in serial intervals.
In addition, serial intervals do not account for asymptomatic transmission, adding further uncertainty to inferences of speed and strength \citep{park2020time}.

Since the beginning of the SARS-CoV-2 pandemic, the importance of understanding the role of generation-interval distributions in its spread has been constantly emphasized \citep{doi:10.1098/rsif.2020.0144}.
Yet, we still feel that generation intervals are under-appreciated, often viewed as a leeway to estimating $\RR$.
Future studies should consider the possibility that the new variant might have longer generation intervals and evaluate how assumptions about generation intervals affect their conclusions.
